---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

``` {r loaddata, echo=TRUE,message=FALSE}
 
library(lubridate)
library(dplyr)
library(ggplot2)
library(scales)

# Make sure activity.zip file is in your R session working directory. 
# Read in the activities file
activities <- read.csv(file=unzip("activity.zip"), stringsAsFactors=FALSE)
 
activities <- activities  %>%  
        mutate(hour = interval %/% 100, minute = interval %% 100)
activities$date_time <- ymd(activities$date)
minute(activities$date_time) <- activities$minute
hour(activities$date_time) <- activities$hour
activities$hr_min <- format(activities$date_time, '%H:%M')

 


 

```
## What is mean total number of steps taken per day?
```{r mean_total_steps, echo=TRUE}

ActivitiesByDay <-  activities[complete.cases(activities),] %>%
        select(steps,date) %>%
        group_by(date)  %>%
        summarise(totalSteps = sum(steps, na.rm=TRUE)) 

# calculate histogram binwidth based on Freedman-Diaconis 
x<- range(ActivitiesByDay$totalSteps)/nclass.FD(ActivitiesByDay$totalSteps)
 
# used this function to select colors -- show_col(brewer_pal()(10))   
ggplot(data=ActivitiesByDay, aes(ActivitiesByDay$totalSteps)) + 
        geom_histogram( binwidth=x[2],
                        fill="#6BAED6",
                        col="#08306B",
                       alpha = .5) + 
        labs(title="Histogram of Total Steps Taken per Day") +
        labs(x="Steps", y="Count")
 
meanSteps <- format(mean(ActivitiesByDay$totalSteps))
medianSteps <- median(ActivitiesByDay$totalSteps)
       
 
```
Mean of the total number of steps taken per day: `r meanSteps`  
Median of the total number of steps taken per day: `r medianSteps`  


## What is the average daily activity pattern?

```{r daily activity, echo=TRUE}
# calculate 5-minute interval (x-axis) and the average number of steps taken, averaged across  # # all days (y-axis)
ActivitiesByInterval <-  activities[complete.cases(activities),] %>%
        select(steps,interval,hr_min) %>%
        group_by(interval, hr_min)  %>%
        summarise(AvgSteps = mean(steps, na.rm=TRUE)) 

# create a "POSIXct" "POSIXt" to display interval in hour:min format on time-series plot below
xx <- activities$date_time[activities$date == "2012-10-01"]

 g <- ggplot(data=ActivitiesByInterval, aes(x = xx, y = AvgSteps))
p <- g + geom_line() + scale_x_datetime(labels = date_format("%H:%M"), 
                                        breaks=date_breaks("2 hour")  ) +
        labs(title="Daily Activity Pattern", x="5 minute interval", 
                            y ="average steps")
p 

# Determinew Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

ActivitiesByInterval <- ActivitiesByInterval %>%
        ungroup() %>%
        arrange(desc(AvgSteps))

maxavgsteps <- ActivitiesByInterval[[1,"AvgSteps"]]
maxinterval <- ActivitiesByInterval[[1,"interval"]]
        
 


```
Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps? Interval `r maxinterval` on average across all the days in the dataset, contains the maximum number of steps, `r maxavgsteps`


## Imputing missing values
Strategy to replace steps missing values involved the following:   

* calculate mean step for each interval (missing values removed)  
* replace steps missing values with the corresponding mean(steps) for that interval

```{r missing values, echo=TRUE}

# Calculate the number of missing values.

missingValues <-sum(is.na(activities$steps))

# replace steps NA values

activitiesNoNA <- activities %>%
        group_by(interval) %>%
        mutate(meanrepNA = mean(steps, na.rm=T))
#replace steps NA values with meanrepNA
activitiesNoNA$steps[is.na(activitiesNoNA$steps)] <- activitiesNoNA$meanrepNA

UpdtActivitiesPerDay <-  activitiesNoNA %>%
        select(steps,date) %>%
        group_by(date)  %>%
        summarise(totalSteps = sum(steps)) 

# calculate histogram binwidth based on Freedman-Diaconis
x<- range(UpdtActivitiesPerDay$totalSteps)/nclass.FD(UpdtActivitiesPerDay$totalSteps)

ggplot(data=UpdtActivitiesPerDay, aes(UpdtActivitiesPerDay$totalSteps)) + 
        geom_histogram( binwidth=x[2],
                       col="#08306B", 
                       fill="#6BAED6", 
                       alpha = .3) + 
        labs(title="Histogram of Total Steps Taken per Day") +
        labs(x="Steps", y="Count")


meanStepsNew <- format(mean(UpdtActivitiesPerDay$totalSteps))
medianStepsNew <- format(median(UpdtActivitiesPerDay$totalSteps))



```
Mean of the total number of steps taken per day: `r meanStepsNew`  
Median of the total number of steps taken per day: `r medianStepsNew`  

## Are there differences in activity patterns between weekdays and weekends?

Step activities is more active during the daytime hours on the weekend days.

```{r weekdays vs weekend, echo=TRUE}

# create a factor variable for weekday and weekend day
activitiesNoNA <- activitiesNoNA %>%
        mutate(day = wday(date, label = TRUE))

activitiesNoNA$typeofday[as.character(activitiesNoNA$day) %in% c("Sat","Sun")] <- "weekend"
activitiesNoNA$typeofday[as.character(activitiesNoNA$day) %in% c("Mon","Tues","Wed","Thurs","Fri")] <- "weekday"
activitiesNoNA$typeofday <- as.factor(activitiesNoNA$typeofday)

# # calculate 5-minute interval (x-axis) and the average number of steps taken, averaged across weekdays and weekends

WeekdayVsWeekend <-  activitiesNoNA %>%
        select(steps,interval,typeofday) %>%
        group_by(typeofday,interval)  %>%
        summarise(AvgSteps = mean(steps, na.rm=TRUE)) 

# create a "POSIXct" "POSIXt" to display interval in hour:min format on time-series plot below
xx <- rep(activitiesNoNA$date_time[activities$date == "2012-10-01"],2)
 
# generate time series plot
g <- ggplot(data=WeekdayVsWeekend, aes(x = xx, y = AvgSteps ))
p <- g + geom_line() +facet_wrap(~typeofday, ncol=1) + 
        scale_x_datetime(labels = date_format("%H:%M"), 
        breaks=date_breaks("2 hour")) + labs(title="Daily Activity Pattern",
                x="5 minute interval",  y ="average steps")
p
```

